---
title: "Basic syntax of the R language"
---

## R code

### Coding in R

R code is organized in *scripts*. Scripts are files ending in `.R` and can 
contain R code and comments (preceded by `#`), e.g. a file `myfile.R` containing

```{r, eval = FALSE}
1 + 1 # an easy summation
```

In RStudio, a new script is created by clicking on `File/New File/R Script` or  
the `+` sign at the top left of the IDE.

### Running R code in RStudio

R code can be run at the console (by typing `Enter`) or from an R script:

* per line or per selection by clicking on `Run` 
* entire R files can be `source`'d, running `source("myfile.R")` either on the 
  command line or from another R script (not covered in these tutorials)


### R commands 

R is an interpreted language, and commands inside a script are parsed line by 
line. Long commands can be split into several lines, but you have to make sure that 
during any line break, the command is still "incomplete".

For example, the following command works perfectly fine: 
since line 1 is incomplete, line 2 is considered part of the command in line 1.

```{r}
2 * 
2
```

In contrast, if `*` is put into the second line, the first line is considered 
complete, and the second command is syntactically wrong, resulting in an error:

```{r, eval = FALSE}
2
* 2
```

```{r, echo = FALSE}
cat("Error: unexpected '*' in '*'")
```

Multi-line commands are often used make complex function calls more easy to read. 
For example, the following command creates a variable of type "data frame" 
(see Tutorial "Data types") with three columns:

```{r}
data.frame(
    x = 1:3,
    y = 4:6,
    z = 7:9
)
```

## Operators

### Assignment operator

To assign a value to a variable, use the `<-` operator:

```{r}
x <- 3
y <- 1 + 1
```

(In fact, there are two assignment operators, `<-` and `=`, but it is recommended 
to always use `<-`, which will be done consistently in this module.)

It is important to write `<-` as a single symbol without blank space (not `< -`), 
otherwise it is interpreted as "smaller than minus".

Assigned values will not be displayed at the console. To display
the value of a variable, type the variable name as a separate command 
(an alternative is to use parentheses `()`, see below)

```{r}
x <- 3
x
```

If you are using RStudio, you can also look at defined variables in the 
"Environment" tab.

### Colon operator

The colon operator `:` creates regular numeric sequences:

```{r}
1:10
1.5:4
5:1
```

### Math operators

The main math operators are 

* `+` (plus)
* `-` (minus)
* `*` (times)
* `/` (divide)
* `^` (power)

All of these operators are vectorized, which means any of the arguments can be
vectors:

```{r}
1:3 + 3:1
```


If the two inputs are vectors of different sizes, the shorter one will 
be automatically *recycled* (i.e., expanded to a vector of the matching size):

```{r}
1:3 + 2
```

### Comparison operators

Comparison of numeric variables results in logical variables (`TRUE`/`FALSE`).
It can be done with operators 

* `==` (equal to; not to be confused with `=`)
* `!=` (different from)
* `>` (greater than)
* `<` (less than)
* `>=` (greater or equal)
* `<=` (less or equal)

Again, all comparison operators are vectorized. See the following exampes:

```{r}
1:3 > 3:1
1:3 != 2
```

### Logical operators

The following logical operators are defined:

* `!` (negation, vectorized)
* `&&` (logical AND)
* `||` (logical OR)
* `&` (logical AND, vectorized)
* `|` (logical OR, vectorized)

See the following examples.

```{r}
!TRUE
TRUE && FALSE
TRUE || FALSE
FALSE || FALSE
```

### Parenthesis and brace operators

The parenthesis operator `(` is used for two purposes:

(1) change operator precedence by grouping

    ```{r}
    1:5+2
    1:(5+2)
    ```
    
    In complex expressions, it is recommended to always make precedence clear via `()`.

(2) make the result of an assignment visible

    ```{r}
    x <- 5+5   # value of x not visible
    (x <- 5+5) # value of x visible 
    ```

The brace operator `{` can be used to group several commands. This is useful in
many contexts, for example in function definitions (see tutorial "Defining a function")
and if/else statements or loops (see tutorial "Control flow"). Only the last 
command inside `{}` is returned:

```{r}
{
    x <- 1+1 #
    x        # not returned
    x+3      # returned
}
```

## Help

To get help about a particular function, use `?` or `help()` (the two produce 
identical results):

```{r}
?sum
help(sum)
```

The help contains information about the function's syntax (i.e., expected arguments
and output format), comments about the purpose and implementation of the function, 
and finally, some examples for how to use the function. 

For beginners, the R help is quite difficult to understand, don't worry if you
feel overwhelmed!

To assess the function help for operators like `+` or `:`, the operator name has 
to be enclosed into backticks:

```{r, eval = FALSE}
?`:`
help(`:`)
```

