---
title: "Control flow in R"
---

This tutorial covers control flow operators:

* `if`/`else` statements
* `for`/`while` loops

## If/else statements

An if/else statement executes code under a certain condition, specified via a boolean
(a variable which is either `TRUE` or `FALSE`). The basic syntax is the following:

```{r, eval = FALSE}
if (boolean) {
    # code to be executed if boolean is TRUE
}
```

or, if some other code shall be executed if `boolean` is `FALSE`, 

```{r, eval = FALSE}
if (boolean) {
    # code to be executed if boolean is TRUE
} else {
    # code to be executed if boolean is FALSE
}
```

Let's give an example:

```{r}
x <- rnorm(1)         # normally distributed pseudorandom number
x
if (x > 0) {
    print("x is positive")
} else {    
    print("x is negative")
}
```

The result of an if statement can also be assigned into a variable. For example,
the above code can be rewritten as follows:

```{r}
text <- if (x > 0) {
    "x is positive"
} else {    
    "x is negative"
}
print(text)
```

If an if statement is used that way without any else-statement and the condition
is not met, the resulting value will be `NULL`.

If/else statements can also be chained, covering more than two possible outcomes:
```{r}
if (x > 1) {
    print("x is greater than 1")
} else if (x >= -1) {
    print("x is in [-1,1]")
} else {
    print("x is less than -1")
}
```

A `switch` provides another way of control flow between any number of cases
(not covered in this tutorial).

## Loops

Loops allow to repeat an instruction many times. There are three types of loops
in R:

* `for`-loops repeat an instruction a predefined number of iterations, each time changing 
a single input parameter.
* `while`-loops repeat an instruction as long as a logical condition is satisfied. 
* `repeat`-loops can always be replaced by `while`-loops and should be avoided altogether.

### For-loops

In most cases, a `for`-loops is the appropriate type of loop. The basic syntax 
of a `for`-loop is as follows:

```{r, eval = FALSE}
for (variable in vector) {
    # code to be run
}
```

Most often, the loop is written as

```{r, eval = FALSE}
for (i in 1:n) {
    # code to be run (using variable 'i')
}
```

where the number of iterations `n` is defined before in the script and where `i` 
is called the running index. 

The following example computes the first ten square numbers (note the allocation
of vector `x` before the loop):

```{r}
n <- 10
x <- numeric(n)
for (i in 1:n) {
    x[i] <- i^2
}
x
```

In fact, in this case it is not necessary to use a `for`-loop because the power 
function is vectorized:
```{r}
(1:n)^2
```

However, not all `for`-loops can be replaced in this way, in particular when the
computations inside the loop are more complex, because of:

(1) a non-vectorized computation inside the loop, or
(2) a computation inside the loop depending on something computed 
    in a previous iteration

An alternative for many `for`-loops is the use of the `apply` family of functions
(not covered in this tutorial).

### While-loops

`while`-loops are more general than `for`-loops; in particular, they can be used if 
the number of iterations is not known a priori (for example, it might depend on 
random numbers generated within the loop). 

The basic syntax of a `while`-loop is as follows:

```{r, eval = FALSE}
condition <- TRUE
while (condition) {
    
    # code to be run
    
    condition <- # ... result (TRUE/FALSE) of a logical operator, e.g. ">"
}
```

See the following example of a `while`-loop, counting the number of "failures" 
until the first "success" of a binary outcome with success probability of 10%
(remember that `!` is the negation operator):

```{r}
nfail <- 0
success <- FALSE
while (!success) {
    x <- rbinom(1, size = 1, prob = 0.1)   # see Tutorial "Probability & Sampling"
    success <- x == 1
    if (!success) {
        nfail <- nfail + 1
    }
}
nfail
```



