---
title: "Manipulating data frames"
---

In this tutorial, we will discuss common workflows with data frames:

- obtaining summary information (functions `head`, `summary`)
- adding/removing columns
- subsetting via `subset`
- combining data frames (`cbind`, `rbind`,`merge`)
<!-- switching between "long" and "wide" formats -->
<!-- split/apply/combine paradigm -->

Make sure to be familiar with the `$` operator for data frames (tutorial "Subsetting")

We will take the "birthwt" dataset as an example:

```{r}
birthwt <- read.csv("birthwt.csv")     # see tutorial "Data import/export"
```

## Summarizing a data frame

For a small data frame, you can simply type its name at the console and have a
look at its contents. However, for larger data frames, a more compact summary is 
useful:

1. `head` allows to see the first rows of a data frame (an optional argument 
    allows to specify the number of rows to see, the default being 6):
    
    ```{r}
    head(birthwt)
    ```
    
2. `summary` computes a summary statistic for each column in the data frame. The
    type of summary depends on the type of column (e.g., numeric or factor):
    
    ```{r}
    summary(birthwt)
    ```


## Adding / removing columns

A new column (e.g., a derived variable) can be added by subassigning into a new
column name:

```{r}
birthwt$low <- birthwt$bwt < 2500
names(birthwt)
```

A column can be removed by subassigning `NULL` into the column:

```{r}
birthwt$low <- NULL
names(birthwt)
```

## Subsetting via `subset` 

Often, only part of a dataset (certain rows) is suitable for a certain statistical
analysis. In principle, this could be done using the subsetting operator `[` in a
matrix style (see tutorial "Subsetting"). However, a much more intuitive and concise
syntax is implemented in function `subset`:

For example, to retrieve only the rows with `ftv > 3`:

```{r}
birthwt[birthwt$ftv > 3, ]
subset(birthwt, ftv > 3)
```

Using (vectorized) logical operators (`!` / `&` / `|`), several conditions can 
be easily combined, and the advantage of `subset` becomes more apparent:

```{r}
birthwt[birthwt$ftv > 2 & (birthwt$smoke == "yes" | birthwt$ht == "yes"), ]
subset(birthwt, ftv > 2 & (smoke == "yes" | ht == "yes"))
```


## Combining data frames

Data frames can be combined in different ways:

(1) using `rbind` and `cbind` (similar to matrices, see tutorial "Data types")
    ```{r}
    rbind(data.frame(x=1,y=2),
          data.frame(x=3,y=1))
    cbind(data.frame(x=1:3),
          data.frame(y=2))
    ```
    
(2) through merging using common column(s):

    ```{r}
    df1 <- data.frame(id = 1:3, x = 4:6)
    df2 <- data.frame(id = 1:3, y = 7:9)
    merge(df1,df2)
    ```
    
