---
title: "Data import/export"
---

R offers many functions to import/export data in various formats. In this short 
tutorial, we will restrict ourselves to the `.csv` format (comma separated values), 
and cover

* import of tabular data from a `.csv` file into a data frame
* export of a data frame into a `.csv` file

## Data import

### Specifying a file path

To import data from a `.csv` file, use function `read.csv()`:

```{r}
df <- read.csv("birthwt.csv")   # see tutorial "Manipulating data frames"
summary(df)                     # for more information about this dataset 
```

The file can be specified using either an *absolute path* or a *relative path* 
with respect to your current working directory. Above, a relative path was used,
we can see that the dataset was indeed in this folder:

```{r}
getwd()
list.files(getwd(), pattern = "*.csv")
```

You can also use the absolute path, or set the working directory via function 
`setwd()`. Note that on Windows, you have to replace `\` in path names by `/`.

For convenience, there is also a function called `read.csv2()`, which uses 
different defaults for separators (`,` as decimal point and `;` as delimiter).

### Formatting options

Beyond the file to import, `read.csv` offers a large number of options to 
customize the import. The most important are listed below:

- `header`: if `TRUE`, uses the first row of the `.csv` file as column names. In
  most cases, this is what you want, otherwise set to `FALSE` and provide column 
  names via argument `col.names`.
- `stringsAsFactors`: if `TRUE`, all character columns are treated as factors, 
  rather than character vectors. Again, this is often appropriate, but some 
  characters like ID labels are not appropriately treated as factors. Hence, 
  some post processing might be required. With R version 4.0.0, the default for 
  `stringsAsFactors` changed from `TRUE` to `FALSE`.

## Data export

Function `write.csv()` writes a data frame to a `.csv` file, and again, 
absolute or relative paths can be used:

```{r, eval = FALSE}
write.csv(df, file = "birthwt2.csv")
```

This function offers much less options than `read.csv()`, the most important being

- `row.names`: if `TRUE` (the default), then row names are written as a first
  dataset column. I always use `row.names = FALSE`.

Importantly, in contrast to `read.csv()` it is **not possible to customize** the 
following arguments:

- `col.names`: column names are always taken from the data frame itself
- `dec`: the decimal point is always `.`
- `sep`: the delimiter is always `,` 

To use (`,`/`;`) for `dec`/`sep` instead of (`.`/`,`), you must use `write.csv2()`.


