---
title: "Data types"
---

This tutorial covers the fundamental data types in R:

* Vectors
* Factors
* Matrices
* Data frames
* Lists
* Functions

## Vectors

A vector contains one or several values of a particular type. The most important
types are:

- **numeric** (e.g., `1.41434`)
- **integer** (e.g., `5` or `-2`)
- **logical** (`TRUE` or `FALSE`)
- **character** (e.g., `"tutorial"`, `"peanut"`, ...)

In addition, each type has a special value `NA`, indicating missing values.

Vectors of a certain length can be allocated (initialized) as follows:

```{r}
numeric(4)
integer(4)
logical(3)
character(6)
```

Vectors can be interconverted between the different formats:

```{r}
as.character(c(2.34,1.42,0))
as.logical(c(2.34,1.42,0))
as.numeric(c(TRUE,FALSE))
as.character(c(TRUE,FALSE))
```


In most cases, the distinction between *numeric* and *integer* types is not very
important, since they are almost always automatically converted to the required format.

Values/vectors can be combined using the command `c`:

```{r}
c(1, numeric(2), 2.3)
c("a", "b", "c")
c(TRUE, TRUE, FALSE)
```

Regular sequences of numeric vectors can be created in different ways, too:

(1) the `:` command:

```{r}
1:4
2.5:4.5
```

(2) function `seq()`:

```{r}
seq(from = 1, to = 7, by = 2)
```

Vectors can also be *named*. Then, each element of the vector has its own name:

```{r}
c(1,2)     # unnamed values
c(a=1,b=2) # values named "a" and "b"
```

## Factors

The data type "factor" is designed for categorical variables, i.e. variables with
a pre-defined set of values (usually, the number of categories is much smaller 
than the length of the vector).

```{r}
x <- factor(c("small","medium","small","big"))
x
```

The possible values are called "levels".

```{r}
levels(x)
```

It is also possible for a factor to have a level which is not used in the vector
itself (for example, this allows to document that a certain category was not 
observed). In this case, define the levels as an input argument to `factor`:

```{r}
y <- factor(c("small","medium","small"), levels = c("small","medium","big"))
y
```

Factors can be *unordered* (the default) or *ordered* -- corresponding to nominal and 
ordinal data types. For example, variable `y` above could be encoded as ordered.
In this case, always use argument `levels`, and in increasing order:

```{r}
y <- factor(c("small","medium","small"), levels = c("small","medium","big"), ordered = TRUE)
y
```
Ordered factors can be compared using comparison operators (see Tutorial 
"Basic syntax"), quantiles can be determined, and many plotting functions respect
the ordering during display.

For efficiency, factors are implemented on top of integer vectors with the levels
as display values. This has important consequences for manipulation of factors 
because many functions trigger a conversion of a factor to its underlying type 
"integer", which is generally not what you want. 
For example, concatenation first converts `y` to integer, then to character, 
to produce a vector consistent with the second argument:

```{r}
c(y, "small")
```

## Matrices

Matrices can be created using function `matrix()`. Arguments `nrow`/`ncol` 
specify the size of the matrix, and one of them can be left blank.

```{r}
matrix(1:6, nrow = 3)
```

For matrices, the operator `*` denotes *element-wise multiplication*. For 
*matrix multiplication* (which is what you often want), use `%*%` instead of `*`:

```{r}
M <- matrix(1:4, nrow = 2)
M
x <- matrix(1:2, nrow = 2)
x
M %*% x
```

Invertible (square) matrices can be inverted using function `solve`:

```{r}
solve(M)
```

Several vectors/matrices can also be combined by row or column using functions
`rbind` and `cbind`:

```{r}
rbind(1:3,4:6)  # arguments interpreted as row vectors
cbind(1:3,4:6)  # arguments interpreted as column vectors
cbind(M,M)      # returns matrix (M M)
                # returns matrix (M)
rbind(M,M)      #                (M)
```

## Data frames

Data frames are designed to hold data. It is a tabular format, similar to a 
matrix, but where different columns can have different types 
(e.g. numeric or character). In many cases, each row corresponds to a different 
observation/record, whereas each column corresponds to one observable quantity.

Data frames can be defined as follows:

```{r}
t  <- c(1, 2, 3)
id  <- c("a", "b", "c")
df <- data.frame(time = t, ids = id)
df
```

For unnamed arguments, `data.frame()` uses the argument names as column names:
```{r}
R  <- c(1, 2, 3)
df <- data.frame(R)
df
```

Most of the time, data frames will be read from a tabular data file (e.g., 
a .csv file) -- see Tutorial "Data import/export". Often, data frames need 
further post-processing like subsetting. This is covered
in Tutorial "Manipulating data frames".

## Lists

A list is like a vector, but it allows to hold heterogeneous data. Like vectors,
lists can be unnamed or named, which depends on whether argument names are provided
to the `list` function (most often, named lists are used):

```{r}
list(1:3,
     "A",
     data.frame(r=1:3)
)
list(x = 1:3,
     y = "A",
     z = data.frame(r=1:3)
)
```

Lists can also be nested:

```{r}
list(x = 1:3,
     y = list(z = 1)
)
```

## Functions

Here a very short introduction to functions is given. Further details about 
functions are provided in the tutorials "Function calls" and "Defining a function".

Each function consist of three parts: 

* the formal arguments (including argument names and optionally default values)
* the function body (the actual function content)
* an environment in which it was defined (see Tutorial "Defining a function")

All R functions return exactly one value. More complex output must be provided
in a suitable data format. Often, this is a list.

R features *first-class functions*. This means that functions are a data type as
any other one:

* functions can be passed as arguments to other functions
* functions can be stored in lists
* functions can be modified (advanced)

Typing a function name at the console reveals 
the content of a function. For almost all base R functions, the output is 
extremely difficult to read. Here's an easy example, function `force`, 
which just evaluates its input.

```{r}
force
```

Here, `x` is the single formal argument, `x` is also the function body, and the
function environment is the (namespace of the) "base" package.
