---
title: "Function calls"
---

R functions can be called in many different ways. In this tutorial, the logic of 
argument processing is described.

## Formal vs. input arguments

All arguments defined in R functions have names, which can be seen in the function help
(check this for the examples to follow). They are called *formal arguments*, to
distinguish them from the *input arguments*, which are used in a function call.

When calling a function, it is optional to provide the name of a formal argument. 
For example, function `sqrt` (square root) has one formal argument, `x`, and 
the following calls are both valid:

```{r}
sqrt(4)
sqrt(x=4)
```

If unnamed arguments are provided as inputs to a function with several formal 
arguments, these are matched *by position*:

```{r}
rep.int(2.3, 4)
```

In contrast, named arguments provided to a function with several formal arguments 
are matched *by name*:


```{r}
rep.int(x = 2.3, times = 4)
rep.int(times = 4, x = 2.3) # same as above
```

Named and unnamed input arguments can be mixed. In this case, a two-step processing is 
done:

* first, all named input arguments are matched by name
* then, all unnamed input arguments are matched by position to the *remaining* 
  formal arguments

Let's consider the above example again:
```{r}
rep.int(times = 4, 2.3) # same as above
```

This function call syntax is extremely flexible, but it can also be quite confusing 
(also for yourself!). The advantage of using unnamed input arguments is that the 
function call becomes more concise. However, it is generally recommended to use 
only a very limited number of unnamed input arguments, and to only use these for
the first arguments.  Moreover, it is recommended to always provide all input 
arguments in the order of the formal arguments. Some people even recommend using
named input arguments exclusively.

## Default argument values

Most R functions define default values for some of their (formal) arguments. As an 
extreme example, function `read.table` has 25 formal arguments, but only one of 
them is mandatory (the `file` from which to read) -- all the others specify 
settings on how to read the information in `file`. 

The logic of argument matching described above is done independently of whether 
a default is defined for a formal argument. However, for the formal arguments 
with default values, it is optional to provide them as input arguments.

For example, function `log` has two formal arguments, `x` and `base`. Argument 
`x` is mandatory, but `base` has the default `exp(1)` (natural logarithm).

```{r}
log(8)                  # natural log
log(8, base = exp(1))   # same as above
log(8, base = 2)        # 2^3 = 8
```

##  The argument `...`

There is a special formal argument called `...`. This argument can match any 
number of input arguments that are not matched by other formal arguments.

For example, function `c` has a `...` argument, and hence, it can accept any 
number of inputs:

```{r}
c(1)
c(1,2,3)
c(234,553)
```

Argument `...` can match both named and unnamed inputs. For example, `c` uses
named arguments to create a named vector. 

Care must be taken to avoid matching another formal argument accidentally. 
For example, function `data.frame` has a `...` argument. Now imagine you want 
to create a data frame with two columns `row.names` and `column.names`:

```{r}
data.frame(row.names    = c("A","B","C"), 
           column.names = c("a","b","c"))
```

The result is not as desired because `row.names` is a formal argument of function 
`data.frame`.

## Non-standard evaluation

If you know another programming language, you will most likely be surprised by 
a feature called *non-standard evaluation*: R functions can access not only the 
*values* of the input arguments passed to them, but also the *commands* that 
resulted in these values.

For example, the axis labels in the following plot are produced by non-standard 
evaluation:

```{r}
x <- 1:10   # creates the vector [1 2 3 ... 10]
plot(x, x^2)
```

However, non-standard evaluation can do much more than just creating labels. For 
example, the `subset` function (see tutorial "Manipulating data frames" for 
more details) uses non-standard evaluation to evaluate the input argument in a 
different context. The following call to `subset` evaluates `x` and `y` not 
as a variable in the workspace, but as a *column* in the first argument, `df`:

```{r}
df <- data.frame(x = 1:3, y = 4:6)
df
subset(df, x > 1 & y < 6)
```

Therefore, the above call actually means

```{r, eval = FALSE}
subset(df, df$x > 1 & df$y < 6)
```

Non-standard evaluation yourself can become quite complicated and a detailed 
description is well beyond the scope of this tutorial, but it is important 
to be aware that such as feature exists.



