---
title: "R packages"
---

The core R installation contains fundamental tools for statistical data analysis. 
More advanced techniques, as well as other capabilities beyond classical statistical 
modelling, are contained in **R packages**. 

## Pre-installed packages

In fact, the base R installation is already organized in packages, namely
`base`, `methods`, `datasets`, `utils`, `grDevices`, `graphics`, and `stats`.

We can see this by looking at the search path:

```{r, echo = FALSE}
library("deSolve")
detach("package:deSolve")
```

```{r}
search()
```

In addition to the above mentioned packages, there are a few additional elements 
on the search path which are not important for us now.

## Install and load a package

Many R packages are hosted on CRAN (comprehensive R archive network). To install
a CRAN package use function `install.packages()`:

```{r, eval = FALSE}
install.packages("ggplot2")
```

Once a package is installed, there are two ways of using functions from this 
package: 

(1) via the scoping operator `::`, using the syntax `package::function()`, e.g.

```{r, eval = FALSE}
ggplot2::ggplot()
```

This is an unambiguous way of calling a function from a certain package, but the
command is a bit long. Therefore, an alternative is

(2) by loading the package (this is what you will usually do)

```{r}
library("ggplot2")
```

This call makes all functions from the package available without any need to use
the scoping operator, i.e. we can now use

```{r, eval = FALSE}
ggplot()
```

We can also see that the `ggplot2` packages was added to the search path:

```{r}
search()
```

However, conflicts can occur if two loaded packages define a function
with the same name. In this case, a warning is displayed and only the function 
defined in the package loaded most recently can be called without the scoping 
operator.

## Important packages

Below follows a list of useful packages for this module, along with their 
capabilities.

Package   | Functionality
----------|--------------
`deSolve` | ODE solving capacities (required)
`ggplot2` | Flexible graphical system
`MASS`    | Example datasets
`reshape2`| Manipulation of data frames 




