---
title: "Plotting in R"
---

R users have the choice between different plotting frameworks:

* base R plots 
* package `ggplot2`
* other packages like `lattice`

This tutorial covers the main syntax for creating R plots using either base R
or package `ggplot2`.

## Plotting in base R

Base R offers a large number of plotting functions (e.g. `plot`, `hist`, 
`barplot`, `boxplot`, ...). The choice of an adequate plotting function for a 
particular graphical representation in descriptive statistics is covered in
"Task 1: Descriptive statistics". Here, the common syntax of plotting functions 
is described, using the `plot` function as an example.

### Arguments to plotting functions

Calling function `plot` with two arguments (x and y) already suffices to generate
a first plot:

```{r}
x <- 1:10
y <- x^2
plot(x,y)
```

The plotting functions have a number of optional arguments customizing the appearance
of the plot. The most important ones are:

* `type` (`"p"` for points, `"l"` for lines, ...)
* `main`: plot title
* `xlab`/`ylab`: axis labels
* `xlim`/`ylim`: axis limits (provided as a two-element vector, like `c(0,10)`)
* `log`: which axes to represent on log-scale, `""` (none, default)/`"x"`/`"y"`/`"xy"`

Let's give the above plot a better appearance:

```{r}
plot(x, y, 
     type = "l",
     ylab = "f(x) = x^2",
     main = "Illustration of a quadratic function")
```

For a complete list of options to the `plot` function, see `?plot.default`.

### Modifying plots

To add graphical elements to an existing plot, you can use functions `points` 
and `lines`:

```{r}
plot(x, y)
lines(x, y+3)
lines(x, y-3)
```

You always have to initialize a plot before using `points`/`lines`. The plot 
can also be initialized blank, but in this case you need to specify `xlim` and
`ylim` values a priori (which is anyways a good idea if you want to draw several elements):

```{r}
plot(NA, xlim = c(1,10), ylim = c(0,100), xlab = "x", ylab = "f(x) = x^2")
lines(x = x, y = y)
points(x = x, y = y)
```

### Graphical parameters

In addition to the arguments to plotting functions, there are also so-called 
*graphical parameters*, defined in function `par`. Some of these can be set 
inside the plotting functions themselves and are passed on to `par`, others can
only be defined by calling `par`. The list of options is very long and the 
details are quite technical. Below ist a list of the most important options,
which can also be provided via plotting functions themselves (so you won't need
to call `par` for these):

* `lty`: line type; 1=solid (default), 2=dashed, 3=dotted, 4=dotdash, ...
* `pch`: plotting symbol (an integer between 0 and 25)
* `col`: color, specified by one of the following:
    * a string like `"red"` or `"blue"` (`colors()` returns all 657 possible values)
    * an integer specifying a color inside the current color palette.
  
## Plotting with `ggplot`

Package `ggplot2` is an implementation of the "Grammar of Graphics", a theory for
describing the syntax of graphics. It is one of the most widely of all R packages.

### The `ggplot` syntax

Here, a brief overview on the syntax of the `ggplot2` package will be given. There
are a lot of useful online resources for further information; a good starting point is 
[https://ggplot2.tidyverse.org/](https://ggplot2.tidyverse.org/).

Before starting, install (if not yet done) and load the package:

```{r, eval = FALSE}
install.packages("ggplot2")
```

```{r}
library("ggplot2")
```

A ggplot consists of at least three parts:

* a call to `ggplot` (without the "2" in the package name), defining the data to be plotted
* the definition of *asthetics* via function `aes` 
  (which columns in the data frame to represent, and how)
* one or more "geoms" like `geom_point` (which geometrical object to draw)

Let's recreate the simple plot of the quadratic function:

```{r}
x <- 1:10
y <- x^2
df <- data.frame(x = x, y = y)
ggplot(df, aes(x = x, y = y)) + geom_point()
```

In contrast to the base R plotting functions, which directly produce a plot, 
`ggplot` by itself produces an *object* which is rendered as a plot when 
the object is *printed*. This means that the ggplot objects themselves can be 
modified, created in several steps, etc. before the actual plotting. For example,
the above plot can be created stepwise and then printed by typing the variable name:

```{r}
ggobj <- ggplot(df, aes(x = x, y = y))
ggobj <- ggobj + geom_point()
ggobj
```

### Advantages of `ggplot` over base R

The main advantage of `ggplot2` over base R plots is the built-in support for common 
high-level graphical operations, like the following:

1. color the plot per level of a factor
2. draw one line per level of a factor (grouping)
3. create a subplot per level of a factor (called "facetting")

This allows to considerably shorten and simplify the commands needed to create 
such plots.
Since pharmacometric plots are often quite complex, this feature makes `ggplot2` 
well suited for this application.

As an example, let us consider again the `birthwt` dataset (from Tutorials 
"Data import/export" and "Manipulating data frames") and make a complex plot:

```{r}
birthwt <- read.csv("birthwt.csv") 
ggplot(birthwt, aes(x = age, y = bwt, color = smoke, shape = ht))  + 
    geom_point() + 
    facet_wrap("race")
```

In base R, creating such a plot would require at least 20 lines of complex code
involving loops, low-level graphical parameters, data subsetting, etc.

### Disadvantages of `ggplot` over base R

For very simple plots, base R functions provide a more intuitive interface. 
For example, compare the following commands:

```{r}
x <- rnorm(1000)     # 1000 normally distributed random numbers
hist(x)
ggplot(data.frame(x), aes(x = x)) + geom_histogram()
```

At the other end of the complexity scale, base R graphics offer a slightly larger
variety of plots than the `ggplot2` package (although the coverage is quite extensive). 
This very uncommon case will not be discussed in this tutorial.

## Graphical devices

If you're using RStudio, the generated plots will appear in the "Plots" panel.
This is one type of graphical output device, a so-called *screen device*.

The default graphical output device can be seen as follows (if you're using 
RStudio, it should be the RStudio graphical device):

```{r, eval = FALSE}
options('device')
```

One way of creating a file (`.pdf`. or an image format like `.png`) from a plot
is to use the interactive export in the RStudio "Plots" panel. 

Alternatively, you can use a *file device* instead of the screen device, like
`pdf`, `postscript` or `png`.

The syntax for exporting to a file is as follows:
```{r, eval = FALSE}
pdf("My histogram.pdf")       # open the pdf file device
hist(x)
dev.off()   # shut down the device again
```

File devices have further options, like width, height or font style, which you
can look up in the function help.



