---
title: "Probability & Sampling"
---

This tutorial covers the mains R functions describing probability distributions 
and the generation of random samples.

## The `d/p/q/r` functions

In **R**, there is a specific set of four functions for many common probability 
distributions, the *d-p-q-r functions* (for Density -- Probability -- 
Quantile -- Realization):

* `d***`: probability density / probability mass function
* `p***`: (cumulative) distribution function
* `q***`: quantile function (inverse of distribution function)
* `r***`: realization (a sample from the specified distribution)

These functions always have the following syntax:

* the first argument is the same for all probability distributions, however its
  name differs for the four types: `x` for `d***`,  `q` (quantile) for `p***`,
  `p` (probability) for `q***`, and  `n` (sample size) for `r***`
* the remaining arguments are the same for the `d/p/q/r` functions, but differ 
  for each each probability distribution; they are the specific parameters 
  describing the distribution. For some but not all distributions, defaults are 
  defined for the distribution parameters.

We will consider some examples:

**Binomial distribution**: $\mathcal{B}(n,p)$ with $n=10$, $p=0.1$

```{r}
dbinom(x = 1, size = 10, prob = 0.1)   # Probability mass function at x=1
pbinom(q = 1, size = 10, prob = 0.1)   # Distribution function at q=1
qbinom(p = 0.5, size = 10, prob = 0.1) # Quantile function at p=0.5
rbinom(n = 20, size = 10, prob = 0.1)  # Simulate a sample of size 20
```

**Poisson distribution**: $\text{Poiss}(\lambda)$ with $\lambda=2$

```{r, eval = FALSE}
dpois(x = 1, lambda = 2)     # Probability mass function at x=1
ppois(q = 1, lambda = 2)     # Distribution function at q=1
qpois(p = 0.5, lambda = 2)   # Quantile function at p=0.5
rpois(n = 20, lambda = 2)    # Simulate a sample of size 20
```

**Normal distribution**: $\mathcal{N}(\mu,\sigma^2)$ with $\mu=1$, $\sigma=3$

```{r, eval = FALSE}
dnorm(x = 2, mean = 1, sd = 3)    # Probability density function at x=2
pnorm(q = 2, mean = 1, sd = 3)    # Distribution function at q=2
qnorm(p = 0.5, mean = 1, sd = 3)  # Quantile function at p=0.5
rnorm(n = 20, mean = 1, sd = 3)   # Simulate a sample of size 20
```

## Function `sample`

Function `sample` allows to draw a sample from a finite set of values, with or
without replacement:

```{r}
sample(1:10, size = 8, replace = TRUE)
sample(1:10, size = 8, replace = FALSE)
```

The first argument can be any vector:

```{r}
letters
sample(letters, size = 8, replace = FALSE)
```

An optional argument `prob` can be used to specify unequal probabilities to obtain
the values in the sample:

```{r}
sample(1:4, size = 10, replace = TRUE, prob = c(0.7,0.1,0.1,0.1))
```

Function `sample.int` is a faster version for sampling from the set `1:n` 
(among others, this function is very useful for bootstrapping):

```{r}
sample.int(10, size = 10, replace = TRUE)
```


## Random numbers and seeds

The sampling functions `r***` and `sample` use peudorandom numbers based on 
the current processor time. This means that running the same code twice will give
different results if sampling functions are involved:

```{r}
rnorm(10)
rnorm(10)
```

In many cases, this behaviour is desired -- after all, these functions are used 
to describe the outcome of a random process! 
However, there is a reproducibility problem if running the same script twice 
(by different people, at different times, ...) yields different results.
It is therefore strongly recommended to set a *seed* for the random number generator
at the beginning of a script:

```{r}
set.seed(12345) # any number will do
rnorm(10)
```

Then, if the code is run again (here, identical code is used to show this 
behaviour), the results can be reproduced:

```{r}
set.seed(12345) # important: same number as before!
rnorm(10)
```

