---
title: "Subsetting"
---

Subsetting of a variable returns a part of this variable (e.g., a part of a vector).
In this tutorial, you will learn about the three subsetting operators 
in R (`[`, `[[` and `$`). The main difference between them is the following:

- **single brackets** `[`: access one or more elements in a variable 
- **double brackets** `[[`: access the content of a single element in a variable
- **dollar sign** `$`: shorthand notation for `[[` in certain data types

The exact behaviour of these operators depends on (1) the type of subsetting used
and (2) the data type which is subsetted. These dependencies is discussed in more 
detail below. Lastly, subassignment will be discussed.

## Types of subsetting

There are three types of subsetting: numeric subsetting, logical subsetting and 
character subsetting. We will illustrate the difference between these using the 
example of a named numeric vector and the `[` operator.

```{r}
x <- c(two = 2, four = 4, six = 6)
x
```

### Numeric subsetting

This is the most straightforward way of subsetting -- the position of the subset
is provided.

```{r}
x[1:2]
x[1]
```

Negative indices allow to specify indices to be dropped:

```{r}
x[-1]
x[-(2:3)]
```

### Logical subsetting

In logical subsetting, a logical array of the same size as `x` is used to 
specify the subset to be chosen (all `TRUE` elements). The logical array is often
the result of a comparison:

```{r}
x[c(TRUE,TRUE,FALSE)]
x[x < 5]
x[x<3 | x>5]
```

### Character subsetting

If the vector `x` has names, these can also be used for subsetting, for example

```{r}
x['four']
x[c('two','six')]
```

## Subsetting different data types

The choice of subsetting operator depends on the data type which is subsetted. 
We will discuss subsetting of vectors, matrices, lists and data frames. 

### Subsetting vectors

Vectors can be subsetted with the `[` and `[[` operators.

* `[` allows to select one or several elements in a vector, and names are kept.
* `[[` only allows to select a single element in a vector and names are dropped.

Here are some examples where you see this at work.

```{r}
x <- c(2,4,6)
y <- c(two = 2, four = 4, six = 6)
x[1:2]
x[[1]]
y[1:2]
y[[1]]
```

### Subsetting matrices

Matrices are subsetted with the `[` operator (`[[` also works, but doesn't 
provide any benefit). Usually, two indices are specified, one for rows and one 
for columns (there is also the less used linear indexing). Any of the arguments
can be left blank, in which case all rows / columns will be taken.

```{r}
M <- matrix(1:9, nrow = 3)
M
M[1:2,2:3]
M[1:2, ]
```

If the subsetting yields a one-dimensional (row or column) vector, the result
is automatically converted to vector type. To prevent this, use a third argument
`drop = FALSE`:

```{r}
M[2, ] 
M[2, ,drop = FALSE] 
```

### Subsetting lists

Lists can be subsetted with all of the three operators: `[`,`[[` or `$`.

* Subsetting with `[` returns a (subsetted) *list*
* Subsetting with `[[` or `$` returns the *content* of a single  list element.
* Subsetting with `$` is a shorthand notation used for named lists; it can always 
  be replaced by `[[`. The `$` notation is particularly useful for (deeply) nested
  lists.

See the following examples.

```{r}
lst <- list(x = 1:3, y = "abcde", z = TRUE)

lst[1]     # list containing vector x
lst[[1]]   # vector x itself

lst[["x"]]       # character subsetting
lst$x            # equivalent to lst[["x"]], but shorter
```


### Subsetting data frames

Data frames can be subsetted using either the list syntax or the matrix syntax.
We illustrate this for the following toy example.

```{r}
df <- data.frame(x = 1:3, y = c("a","b","c"), z = c(TRUE, FALSE, FALSE))
df
```

List syntax:

```{r}
df[c('x','y')]   # two-column data frame
df['x']          # single-column data frame 
df$x             # content of column x
df[['x']]        # again, identical to df$x
```

Matrix syntax:

```{r}
df[ ,c('x','y')]

df[ ,'x']   # same as df$x
df[ ,c('x','y')]
```

More complex subsetting mechanisms are possible and often used when 
pre-processing data prior to a statistical analysis. This topic is covered in 
the tutorial "Manipulating data frames".

## Subassignments

All of the above mentioned subsetting operations work on the left-hand side of 
an assignment. In this case, the subsetted part of the object is replaced by the
right-hand side of the assignment. 

Here are some examples for subassignments:

(1) Subassignment for numeric vector

```{r}
x <- 6:10
x[2:3] <- 4:5
x
```


(2) Subassignment for a list 

```{r}
x <- list(y = 1, z = "a")
x$y <- 2
x
```

(3) Subassignment for names of a vector

```{r}
x <- c("a"=1,"b"=2)
x
names(x)[1] <- "A"
x
```

