---
title: "Descriptive statistics"
---

<!--JavaScript function for a button to show/hide solutions-->
<script>
function showSolution(exercise) {
    var x = document.getElementById(exercise);
    if (x.style.display === "none") {
        x.style.display = "block";
    } else {
        x.style.display = "none";
    }
}
</script>
 
<!--Main R markdown file-->

We consider a dataset on risk factors associated with low infant birth weight
(modified from the `MASS` **R** package). First, we set the working directory
and read the dataset:

```{r, eval = FALSE}
setwd("~/work/Teaching/Statistics-2021/task-descriptive-stats")
birthwt <- read.csv("birthwt.csv")
```

```{r load-data, echo = FALSE}
birthwt <- read.csv("birthwt.csv", stringsAsFactors = TRUE)
```

The `birthwt` dataset contains the following columns:

* `age`: mother's age in years
* `lwt`: mother's weight in pounds
* `race`: mother's race
* `smoke`: smoking status during pregnancy
* `ht`: history of hypertension
* `ftv`: number of physician visits during the first trimester of pregnancy
* `bwt`: birth weight in grams

```{r summary-stats}
summary(birthwt)
```

**Exercise**  
Determine the scale of measurement for each variable: 

* nominal, 
* ordinal, 
* count (discrete numeric) 
* or continuous?

<!--Solution to exercise 1-->
<button onclick='showSolution("solutionEx1")'>Show/hide solution</button>
<div id="solutionEx1" style = "display:none">
Answer: 

* `age`: continuous
* `lwt`: continuous
* `race`: nominal
* `smoke`: nominal
* `ht`: nominal
* `ftv`: count
* `bwt`: continuous
</div>

------------------------------------------------------------

## Representation of a single variable

### Continuous variable

#### Numerical description

```{r}
bwt <- birthwt$bwt
```

Measures of location
```{r}
mean(bwt)
median(bwt)
```

Measures of dispersion
```{r}
sd(bwt)       # (empirical) standard deviation
var(bwt)      # (empirical) variance
range(bwt)    # c(min(x), max(x))
IQR(bwt)      # interquartile range
```

Sample quantiles
```{r}
quantile(bwt)                       # default: quartiles 
quantile(bwt, probs = c(0.2,0.8))
```

Sample coefficient of variation (measure of dispersion relative to location;
dimensionless). 
```{r}
cv <- sd(bwt)/mean(bwt)
cv
```

Sample mean and sample median both are a measure of location, and 
sample standard deviation and interquartile range both are a measure of dispersion.
The main difference between these measures is that sample mean and sample standard
deviation are more affected by outliers than sample median and interquartile range,
respectively.


#### Graphical description

Continuous variables are usually displayed in a *histogram*, a *boxplot*, 
or a *dot plot* (1D scatter plot).

A **histogram** groups values into predefined intervals (''binning'') and represents
counts in these intervals:
```{r}
hist(bwt)                # default intervals
hist(bwt, breaks = 2)    # coarser intervals
hist(bwt, breaks = 100)  # finer intervals
```

Advantages:

* multimodality and skewness easy to see

Disadvantages:

* 2D representation of 1D data (difficult to group)
* representation depends on interval width

A **boxplot** represents summary statistics depending on sample quantiles:

* median
* quartiles
* ''fences''
    + lower fence: at smallest observation larger than $q_{0.25} - 1.5\cdot IQR$
    + upper fence: at largest observation smaller than $q_{0.75} + 1.5\cdot IQR$ 
* outliers (values outside fences, if any)

```{r}
boxplot(bwt)
```

Advantages:

* 1D representation of 1D data (can be grouped)
* outliers are easily spotted

Disadvantages:

* multimodality not visible
* sample size not visible

A (continuous) *dot plot* shows individual data points along the x-axis, while 
scattering (''jittering'') values along the y-axis to avoid overplotting.
```{r}
stripchart(bwt, method = "jitter")
```

Advantages:

* Shows the actual data!
* ''almost 1D'' representation of 1D data (can be grouped)

Disadvantages:

* only adequate for small datasets

------------------------------------------------------------

### Discrete variable

Generally, a discrete variable is summarized by counting number of datapoints
in each category; this information can be displayed directly as a *table* 
(either absolute or relative frequencies) or 
graphically as a *barplot*.

```{r disp-single-discrete}
race <- birthwt$race
table(race)                 # absolute frequencies
table(race)/length(race)    # relative frequencies
barplot(table(race))
```

* Discrete *numeric* variables (i.e., count data) can be further summarized 
  numerically, similar to continuous variables. In particular we can compute 
  their mean, variance/standard deviation, median (or, more generally, 
  quantiles), minimum/maximum, and interquartile range.
* Rather than using the coefficient of variation, the *index of dispersion* is 
  often computed (ratio of sample variance to sample mean).
  It is 1 for Poisson distributed data; when the dispersion index is $>1$,
  data are called overdispersed and for a dispersion index $<1$, they are called
  underdispersed.
* Count variables with many categories are sometimes displayed as if they were 
  continuous (e.g., via a histogram or boxplot). However, a histogram and a 
  barplot are not the same!

**Exercise**  

Classify the frequency of physician visits according to its dispersion index.

<!--Solution to exercise 2-->
<button onclick='showSolution("solutionEx2")'>Show/hide solution</button>
<div id="solutionEx2" style = "display:none">
Answer:
```{r}
var(birthwt$ftv)/mean(birthwt$ftv)
```
$\Rightarrow$ dispersion index greater than one, hence frequency of physician visits is 
overdispersed.
</div>

------------------------------------------------------------

### Labelling

It is important to use meaningful axis labels and titles, such as in the 
following examples:

```{r}
barplot(table(race), ylab = "Frequency", main = "Race distribution in the birthwt dataset")
hist(bwt, xlab = "Infant birthweight in [g]", main =  "")
```

**Exercise**  
Summarize each variable in the `birthwt` dataset appropriately. Use adequate
x/y labels and a title.

<!--Solution to exercise 3-->
<button onclick='showSolution("solutionEx3")'>Show/hide solution</button>
<div id="solutionEx3" style = "display:none">
Answer:
```{r}
hist(birthwt$age, xlab = "Age of mother in years", main = "")
hist(birthwt$lwt, xlab = "Weight of mother in pounds", main = "")
barplot(table(birthwt$race), xlab = "Mother's race")
barplot(table(birthwt$smoke), xlab = "Did the mother smoke during pregnancy?")
barplot(table(birthwt$ht), xlab = "Mother with history of hypertension?")
barplot(table(birthwt$ftv), xlab = "Frequency of physician visits in first trimester")
hist(birthwt$bwt, xlab = "Birthweight in grams", main = "")
```
</div>

------------------------------------------------------------

## Relationships between two variables

### Both continuous

Two continuous variables are best shown in a *scatterplot*:

```{r}
plot(birthwt$age, birthwt$bwt)
```

The degree of linear association between two continuous variables can be 
expressed by the empirical correlation coefficient $r\in[-1,1]$:

```{r}
cor(birthwt$age, birthwt$bwt)
```

------------------------------------------------------------

### One continuous, one discrete

The values of *non-numeric* discrete variables divide the dataset into different
groups. For each of these groups, summary statistics can be computed. 

For graphical display, boxplots and dot plots are well suited since they can be
placed side by side:

```{r}
boxplot(bwt ~ race)
stripchart(bwt ~ race, method = "jitter")
```

------------------------------------------------------------

### Both discrete

Two discrete variables are best summarized via a *joint frequency table*. To 
avoid potential confusion between variables with the same set of possible 
values, use named arguments:

```{r}
(tab <- table(ht=birthwt$ht, smoke=birthwt$smoke))
```

A graphical representation of 2 discrete variables is less commonly used than a
tabular representation. One possibility is the *mosaic plot*:

```{r}
mosaicplot(tab)
```

Alternatively, a *small multiples* arrangement of *dotcharts* (in base **R**, 
the default labelling is unfortunate for both these possibilities).

```{r}
colnames(tab) <- paste0('smoke: ', colnames(tab))
rownames(tab) <- paste0('ht: ', rownames(tab))
dotchart(tab)
```

... or *stacked/grouped barplots* (this is difficult to do in base **R**, 
therefore one best uses a dedicated **R** package, e.g. `ggplot2`):

```{r, eval = FALSE}
install.packages("ggplot2") #if not done already
```

```{r}
library(ggplot2)
ggplot(birthwt) + geom_bar(aes(x=smoke, fill = ht))
```

**Exercise**  
Represent the following relationships appropriately: 

* `race` and `ht`
* `smoke` and `age`
* `ftv` and `lwt`

<!--Solution to exercise 4-->
<button onclick='showSolution("solutionEx4")'>Show/hide solution</button>
<div id="solutionEx4" style = "display:none">
Answer:
```{r}
table(race=birthwt$race,ht=birthwt$ht)
boxplot(birthwt$age~birthwt$smoke)
plot(birthwt$ftv,birthwt$lwt, xlab = "Number of first trimester visits",
     ylab = "Mother's body weight in pounds")
```
</div>

------------------------------------------------------------

## Data post-processing

### Numerical transformations

Sometimes a variable is transformed numerically (often, *log-transformed*) before it is
represented numerically or graphically. For illustration, we consider 
the `Animals` dataset from the `MASS` **R** package (body and brain size for
different animal species):

```{r, eval = FALSE}
install.packages("MASS") #if not done already
```

```{r}
animals <- MASS::Animals
plot(animals$body,animals$brain)
plot(animals$body,animals$brain, log = "xy")
```

**Exercise**  

Make sure you understand which summary statistics commute with a monotonously 
increasing data transformation like the log-transform, and which ones don't 
(commutation means that if $T(X)$ is a statistic 
and $f$ a data transformation, then $f(T(X)) = T(f(X))$).

<!--Solution to exercise 5-->
<button onclick='showSolution("solutionEx5")'>Show/hide solution</button>
<div id="solutionEx5" style = "display:none">
Answer: 

* mean, standard deviation/variance and range/IQR don't commute with transformation
* all quantile-based summary statistics (median, max/min, quartiles) commute with monotonous transformations
</div>

------------------------------------------------------------

### Subsetting

Only a subset of data might be relevant for a particular analysis. Use function
`subset` to create a sub-dataset (see tutorial "Manipulating data frames").

For example, we might want to consider only the mothers with hypertension:

```{r}
subset(birthwt, ht == "yes")
```

------------------------------------------------------------

### Categorisation

Numeric variables (continuous or count) can be categorized into ordinal variables.
Unfortunately, there is no such function in base R, but the categorization can be
done manually using functions `findInterval` and `factor`: 

```{r}
idx <- findInterval(birthwt$ftv, c(-0.5,0.5,1.5)) 
catftv <- factor(c("never","once","several")[idx], ordered = TRUE)
head(catftv)
```

------------------------------------------------------------

## Further exercises

1)  Represent the following relationships appropriately:  
* `smoke`, `lwt` and `bwt`
* `age` and `bwt` of mothers smoking during pregnancy
* `smoke` and `bwt` of mothers over 30 that never went to a doctor during 
    first trimester of pregnancy

    <!--Solution to exercise E-1-->
    <button onclick='showSolution("solutionExE1")'>Show/hide solution</button>
    <div id="solutionExE1" style = "display:none">
    ```{r}
    plot(birthwt$lwt, birthwt$bwt, col = birthwt$smoke, 
         xlab = "Mother's weight in pounds",
         ylab = "Birth weight in grams",
         main = "Stratification by smoking status")
    legend("bottomright", legend = unique(as.character(birthwt$smoke)), 
           col = unique(as.numeric(birthwt$smoke)), pch = 1)
    ```
    
    ```{r}
    smok <- subset(birthwt, smoke == "yes")
    plot(smok$age, smok$bwt, xlab = "Mother's age in years",
         ylab = "Birth weight in grams",
         main = "Association for smoking mothers")
    ```
    
    ```{r}
    cpx <- subset(birthwt, ftv == 0 & age > 30)
    boxplot(cpx$bwt ~ cpx$smoke, xlab = "Smoking status during pregnancy",
            main = "Mothers over 30 without first trimester visit")
    ```
    </div>

