---
title: "Defining a function"
---

This tutorial covers the basics on how to define an R function yourself. Be sure
to make yourself familiar with the material in tutorial "Function calls" first.

## Function definition

An R function is defined using the following syntax:

```{r, eval = FALSE}
f <- function(arguments) {
    # function body (computations using 'arguments')
}
```

Here, `f` is the name of the defined function, `arguments` specifies the formal 
arguments and the code in the curly brackets `{}` is the function body. 
Command `return` allows to specify the return value.
If `return` is not called, the *value of the last command* is returned by the 
function. *Note: many people use `return` for early returns only.*

Let's start with some basic examples:

(1) A function computing the square of the input:

```{r}
square <- function(x) {
    x^2
}
square(2)
```

(2) A power function with default exponent of 2:

```{r}
power <- function(x, p = 2) {
    x^p
}
power(2)
power(2, p = 3)
```

(3) A function reporting the current time (without any arguments):

```{r}
report_time <- function() {
    str <- date() 
    paste("Current time:", str)
}
report_time()
Sys.sleep(1)    # wait for 1 sec
report_time()
```

(4) A function calling another function on a fixed numeric vector

```{r}
apply_to_123 <- function(f) {
    f(1:3)
}
apply_to_123(mean)
apply_to_123(sum)
```

## Environmental variables

A function can "see" the variables defined in the environment in which it was
defined. Let us first look at the general mechanism:

```{r}
x <- 1
y <- 3
f <- function(x) {
    x+y
}
f(2)
```

Here, variable `x` defined within the function takes priority over variable `x` 
in the environment (i.e., the script). In contrast, `y` is not defined within the
function, hence `f` looks for it in the environment.

Let us now consider two important applications:

(1) Fix some of the input arguments to (non-default) values

Some R functions expect a function as input themselves, for example an ODE 
solver expects a function encoding the right-hand side of the ODE. 

Sometimes, the formal arguments of your function differ from the expected ones.
In this case, a wrapper function can be defined, exploiting environmental variables.

In the following example, `y` is set to the value defined in the script.

```{r}
y <- 3
adder <- function(x, y) x+y
increase_by_y <- function(x) adder(x, y)   # y taken from script
increase_by_y(1)
```

(2) Avoid computationally expensive tasks in a function

If a function is called many times (e.g. when solving an ODE or during 
optimization), you want to avoid computationally complex tasks
like loading data to be repeated many times.

Let's look at the following toy example:
```{r}
f <- function(x) {
    data <- {Sys.sleep(0.1); 4}  # a long task (imagine loading a dataset here)
    data + x
}

# many function calls (imagine an ODE solver / optimizer calling function f)
tstart <- Sys.time()
for (i in 1:10) {
    f(i)
}
tend   <- Sys.time()
tend - tstart         # check the runtime
```

In these cases, it is preferrable to perform the task once outside the function:

```{r}
tstart <- Sys.time()
data <- {Sys.sleep(0.1); 4}
f <- function(x) {
    data + x
}
for (i in 1:10) {
    f(i)
}    
tend   <- Sys.time()
tend - tstart        
```



